Behavior Driven Development in Haskell

### 2012-05-29 paper.tex
Problem wip

Many Haskell programmers tend not to write test code even though the
strong type system of Haskell cannot ensure correct semantics.
Without test code, it is difficult for users to understand usage of
functions and it is difficult for developers to maintain semantics
while progressing software. To motivate Haskell programmers to write
test code, we propose to carry out behavior driven development with
two tools: doctest integrated with QuickCheck for user documentation
and Hspec with combinators extended for developer documentation. With
our method, specification can be used for documentation and automatic
testing.

@@

Unfortunately, the type system of Haskell does not always guarantee
correct semantics.
For example, it does not ensure that division by zero does not occur.
Another example is laws of type classes.
For instance, the {\tt Monad} class has three Monad laws
that its instance \emph{should} satisfy
but
it is the programmers responsibility to ensure that
instance definitions \emph{behave} as \emph{expected}.

@@

\item[evidence]
    Evidence is either an example or a property.  It provides some confidence,
    that a function exhibits a certain behavior.

@@

\begin{quote}\begin{itemize}
    \item
        We want to use properties and examples to provide evidence for
        behavior.
    \item
        Haddock already has markup support for interactive examples,
        and Doctest can verify them.
    \item
        It is not possible to use properties in Haddock comments and
        automatically test them.
\end{itemize}\end{quote}

In the Haskell community, HUnit and QuickCheck are commonly used %% what about SmallCheck?
and the {\tt test-framework} package integrates them.
We can categorize {\tt test-framework} as
a tool for TDD.
Before our work and Hspec, there was no widely used BDD tool for Haskell.

BDD tools used in other programming language communities are
mainly based on example tests,
and do not support property tests.
To our experience is, that it is not difficult to find properties for
pure Haskell functions.
So, when we bring BDD tools to
Haskell community, we should integrate them with properties test.

We can categorize evidence into two types: for users and
and for developers.
Evidence for users are those which are useful as user documentation.
By contrast, evidence for developers are
those which are not useful as user documentation
but should be written as code documentation.
A good example of evidence for developers
is a corner example registered in a bug tracking system.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Problem Summary}

@@

%%%%%%%%%%%%%%%%
\subsubsection{Markup for Examples}

@@

%%%%%%%%%%%%%%%%
\subsubsection{Markup for Properties}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Open problems and related work}

>>>>

Many Haskell programmers tend not to write test code even though the
strong type system of Haskell cannot ensure correct semantics.
Without test code, it is difficult for users to understand usage of
functions and it is difficult for developers to maintain semantics
while extending software. To enhance the value of test code, we
propose to carry out behavior driven development with two tools:
doctest integrated with QuickCheck for user documentation and Hspec
with combinators extended for developer documentation. With our
method, specification can be used for documentation and automatic
testing. This kill-three-birds-with-one-stone aspect would motivate
Haskellers to write test code.

@@

Unfortunately, the type system of Haskell does not always guarantee
correct semantics.
For example, it does not ensure that division by zero does not occur.
Another example is laws of type classes.
For instance, the {\tt Monoid} class has three Monoid laws
that its instance \emph{should} satisfy
but
it is the programmers responsibility to ensure that
instance definitions \emph{behave} as \emph{expected}.

@@

\item[evidence]
    Evidence provides some confidence, that a function exhibits a
    certain behavior.  In Haskell we use examples and properties as
    evidence.

@@

\begin{quote}\begin{itemize}
    \item
        BDD tools for other programming languages use code examples as
        evidence for behavior.
    \item
        When we apply BDD to Haskell we want to use properties and
        examples to provide evidence for behavior.
    \item
        Hspec already has support for both.
    \item
        Haddock has markup support for interactive examples, and
        Doctest can verify them.
    \item
        But it is not possible to use properties in Haddock comments
        and automatically test them.
\end{itemize} \end{quote}

\begin{quote}
Hspec can be used to document the behavior of a system in a way that
is similar to a specification.  All facets of the system are described
with examples and properties, including every possible corner case.

We think that some evidence is useful as user documentation, hence we
want to be able to include it in Haddock comments.  But there is also
evidence that is not useful as user documentation.  E.g. that a
particular \texttt{Monoid} instance is a monoid in the mathematical
sense is self-evident.  So we do not want to include any examples or
behaviors for that into user documentation, but we still want to be
able to automatically test it.

We propose to use Doctest to test evidence that is useful as user
documentation, and Hspec for evidence that is not useful as user
documentation.
\end{quote}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Problem summary}

@@

%%%%%%%%%%%%%%%%
\subsubsection{Markup for examples}

@@

%%%%%%%%%%%%%%%%
\subsubsection{Markup for properties}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Open problems} %% Future work?

\subsection{Reusable spec}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related work}
